## Week 4

### Introduction

Today's exercise exercise will focus on different techniques of clustering and classification. I will use data on housing in areas of Boston and mostly focus on the crime rate in the city. The data can be accessed through the R library [_MASS_](https://cran.r-project.org/web/packages/MASS/index.html). The data contains area-level information on the characteristics of homes (size, value etc.), the demographic composition of the area as well as several variables related to environmental and infrastructural factors. More information on the data is available [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html) and in the original study by [Harrison & Rubinfeld (1978)](https://www.sciencedirect.com/science/article/abs/pii/0095069678900062).

BTW, I decided to hide my code chunks by default in this course diary as they make reading a bit tedious. You should be able to get the codes by clicking the Code-button next to results.

### Overview of the dataset

```{r,warning=F,message=F}
library(MASS)
library(tidyverse)
library(GGally)
library(corrplot)
data(Boston)
dim(Boston)
str(Boston)
summary(Boston)

hist(Boston$crim)

```

The data includes 14 variables from some 500 regions in Boston. All of them are numeric.My main varibale of interest, crime rate, seems highly skewed, with most of the areas having low rates of crime and a have expressing higher rates.

```{r,fig.cap='Correlation plot'}

p.values.mat <-cor.mtest(Boston,
                         conf.level = .95)
cor.mat <- cor(Boston)
corrplot.mixed(cor.mat,
         lower.col='black',
               upper='color',
               tl.col='black',
               tl.cex=0.5,
               number.cex=0.5,
               p.mat=p.values.mat$p,
               sig.level=0.05)
```

For a graphical overview of data, I am using the correlation plot to make the plot to some extent readable (as compared to _pairs_ or _ggpairs_. In the plot, I have crossed out all the correlations not significant at 95% confidence level. Accordingly, it seems that the variable chas is not significantly correlated with most of the variables (probably as it is binary). Most of the other variables are, and there seems to be relatively strong correlations, for instance rad tax 0.91, age and dis -0.75, nox indus 0.76. Most of the correlations seem moderate, between 0.3 and 0.6.Rad and tax for crime rate.

### Linear Discriminant analysis

```{r}
#Scale boston data

boston_scaled <- as.data.frame(scale(Boston))
summary(boston_scaled)
```

Tähän sössöä

```{r}
#Save categories of crime rate
bins <- quantile(boston_scaled$crim)
#Create new crime variable
crime <- cut(boston_scaled$crim,
             breaks=bins,
             include.lowest=T,
             label=c(
               "low",
               "med_low",
               "med_high",
               "high"))
boston_scaled$crim <- NULL
boston_scaled$crime <- crime
#Divide data into test and training sets
n <- nrow(boston_scaled)
#Randomly sample 80% of the original rows
#These are used for training
ind <- sample(n, size=n*0.8)

#Train set
train <- boston_scaled[ind,]
#Test set
test <- boston_scaled[-ind,]

#Correct classes in the test set
correct <- test$crime

#Drop crime from test
test <- dplyr::select(test, -crime)

```

```{r}
#LDA model
lda.fit <- 
  lda(crime~., data=train)

lda.fit


# the function for lda biplot arrows
#(Stolen from Datacamp)
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2,
     col=classes,
     pch=classes)
lda.arrows(lda.fit, myscale = 2)


# predict classes with test data
lda.pred <- predict(lda.fit,
                    newdata = test)

# cross tabulate the results
table(correct = correct, 
      predicted = lda.pred$class)

table(correct = correct, 
      predicted = lda.pred$class) %>%
  prop.table(2) %>% round(digits=2)


```

### K-means clustering

```{r}
#Reload boston
data(Boston)
boston_scaled <-
  as.data.frame(scale(Boston))

#Calculate distances between observations
#I use Euclidean for no specific reason
#except that the km algorithm uses it
#by default

distances <- dist(boston_scaled)
summary(distances)

#Identify correct number of clusters
#Use the WCSS for this purpose

k_max <- 15 #Arbitrary number

twcss <-
  sapply(1:k_max,
         function(k){
           kmeans(
             boston_scaled,k)$tot.withinss})
plot(x=1:k_max,y=twcss,type='l')

#2 seems appropriate
#Run k-means algorithm

km <- kmeans(boston_scaled,centers=2)

#Plot the data set in three parts
pairs(boston_scaled[
  c(1,5,6,7,8,12,13,14)], 
        col=km$cluster)

```

### K-means + LDA

```{r}

km2 <- kmeans(boston_scaled,centers=3)
boston_scaled$km_clust <- km2$cluster

#LDA model (rename to avoid confusion
#in the next step)
lda.fit2 <- 
  lda(km_clust~., data=boston_scaled)

lda.fit2

classes <- 
  as.numeric(boston_scaled$km_clust)

# plot the lda results
plot(lda.fit2, dimen = 2,
     col=classes,
     pch=classes)
lda.arrows(lda.fit2, myscale = 2)


```
```{r,warning=F,message=F}
model_predictors <- 
  dplyr::select(train, -crime)

# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)

# matrix multiplication
matrix_product <- 
  as.matrix(model_predictors) %*% lda.fit$scaling

matrix_product <- as.data.frame(matrix_product)

library(plotly)

plot_ly(x = matrix_product$LD1,
        y = matrix_product$LD2,
        z = matrix_product$LD3,
        type= 'scatter3d',
        mode='markers',
        color=train$crime)

#let's fit still another k-means
#For some reason, the code does not work
#with the original train data
#However, this just loads the boston
#and limirts the same areas so no 
#errors here, I guess

data(Boston)
boston_scaled <- as.data.frame(scale(Boston))
train <- boston_scaled[ind,]
km3 <- kmeans(train, centers=2)

plot_ly(x = matrix_product$LD1,
        y = matrix_product$LD2,
        z = matrix_product$LD3,
        type= 'scatter3d',
        mode='markers',
        color=km3$cluster)

```

### References
Harrison, D. & Rubinfeld, D.L. Hedonic housing prices and the demand for clean air. 1978. Journal of Environmental Economics and Management 5(1), 81-102.
