## Week 6

### Introduction

In this exercise, I am repeating the analyses from the chapters 8 and 9 of the book Multivariate Analysis for the Behavioral Sciences (Vehkalahti & Everitt 2019). In chapter 8, longitudinal analysis on scores of brief psychiatric rating scale among 40 males is conducted. Chapter 9 of the book focuses on the time evolution of weights of rats. Instead of replicating the same analysis with the same data, I will conduct the analyses in chapter 8 with the data on rats from chapter 9 and vice versa.

### Longitudinal analysis of rat growth

First, let's have a quick look at the data on the weight of rats:

```{r}
setwd("~/IODS-project/data")
library(openxlsx)
library(knitr)
library(tidyverse)
library(kableExtra)

rats <- read.xlsx("rats.xlsx")

#Transform vars
rats$id <- as.factor(rats$id)
rats$group <- as.factor(rats$group)
rats$time <- as.integer(rats$time)

#First 10 rows
kable(head(rats,n=10)) %>%
  kable_styling(full_width=F)

str(rats)

table(rats$id)
table(rats$group)
table(rats$time)

```

Selitä jutut

```{r}

labels <- c("1"="Group 1",
            "2"="Group 2",
            "3"= "Group 3")


rats %>%
  ggplot(aes(x=time,
             y=weight)) +
  facet_wrap(~group,
             labeller=labeller(
               group=labels)) +
  geom_line(size=1,aes(col=id)) +
  scale_colour_viridis_d(option='inferno') +
  theme_minimal() +
  theme(legend.position='none') 
  

```

Tracking: standardize values

```{r}

rats <- 
  rats %>%
  group_by(time) %>%
  mutate(mean_weight=mean(weight),
         sd_weight=sd(weight)) %>%
  mutate(std_weight=
           (weight-mean_weight)/
           sd(weight)) %>%
  ungroup()

rats %>%
  ggplot(aes(x=time,
             y=std_weight)) +
  facet_wrap(~group,
             labeller=labeller(
               group=labels)) +
  geom_line(size=1,aes(col=id)) +
  scale_colour_viridis_d(option='inferno') +
  theme_minimal() +
  theme(legend.position='none') 
  

```

mean profiles

```{r}

rats_s <-
  rats %>%
  group_by(group,time) %>%
  mutate(mean=mean(weight),
         sd=sd(weight),
         n=n()) %>%
  mutate(error=
           qt(0.975,df=n-1)*sd/sqrt(n)) %>%
  mutate(lower=mean-error,upper=mean+error)
  
rats_s %>%
  ggplot(aes(x=time,
             y=mean,
             col=group)) +
  geom_ribbon(aes(
    ymax=upper,
    ymin=lower,
    fill=group),
    alpha=0.3) +
  geom_line(size=1) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() + 
  theme_minimal() +
  theme(legend.position='bottom') 

```
```{r}

rats %>%
  ggplot(aes(x=as.factor(time),
             y=weight,
             col=group)) +
  geom_boxplot() +
  #geom_jitter() +
  scale_fill_viridis_d() + 
  scale_colour_viridis_d() +
  theme_minimal() +
  theme(legend.position='bottom') 

```
Summary measure approach

```{r}
rats_s_2 <-
  rats %>%
  filter(time>1) %>%
  group_by(group,id) %>%
  mutate(mean=mean(weight)) %>%
  ungroup()

#filter leaves only one obs into the data
rats_s_2 %>% filter(time==8) %>%
  ggplot(aes(x=group,y=mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", 
               geom = "point",
               shape=23, size=4,
               fill = "white") +
  scale_y_continuous(
    name = 'mean weight during follow-up')

rats_s_2 %>% 
  filter(time==8) %>%
  select(c('id','group','mean')) %>%
  kable() %>% kable_styling(full_width=F)

```

Three outliers

```{r, warning=F}
rats_s_2 %>% filter(id!=2) %>% #group 1
  filter(id!=12) %>% #group 2
  filter(id!=13) %>%
  filter(time==8) %>% #leave only one obs
  ggplot(aes(x=group,y=mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", 
               geom = "point",
               shape=23, size=4,
               fill = "white") +
  scale_y_continuous(
    name = 'mean weight during follow-up')

#Check the previous plot without these outliers

rats_s_t <-
  rats %>%
  filter(id!=2) %>% #group 1
  filter(id!=12) %>% #group 2
  filter(id!=13) %>% #group 3
  group_by(group,time) %>%
  mutate(mean=mean(weight),
         sd=sd(weight),
         n=n()) %>%
  mutate(error=
           qt(0.975,df=n-1)*sd/sqrt(n)) %>%
  mutate(lower=mean-error,upper=mean+error)
  
  
rats_s_t %>%
  ggplot(aes(x=time,
             y=mean,
             col=group)) +
  geom_ribbon(aes(
    ymax=upper,
    ymin=lower,
    fill=group),
    alpha=0.3) +
  geom_line(size=1) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() + 
  theme_minimal() +
  theme(legend.position='bottom') 


```

```{r}
rats_no_out <- 
  rats_s_2 %>% filter(id!=2) %>% #group 1
  filter(id!=12) %>% #group 2
  filter(id!=13) %>% 
  filter(time==8) #leave only one row

anova(lm(mean~group, data=rats_no_out))

#baseline weight
baselines <- 
  rats %>% filter(time==1) %>%
  select(c('id','weight')) %>%
  rename(baseline=weight)

rats_no_out_2 <-
  inner_join(rats_no_out,baselines,by='id')

anova(lm(mean~group + baseline,
         data=rats_no_out_2))

summary(lm(mean~group + baseline,
         data=rats_no_out_2))

```

### Linear mixed effects model

the above analysis is boring

```{r}
#read data
setwd("~/IODS-project/data")
bprs <- read.xlsx("bprs.xlsx")
head(bprs)
#Transform vars

bprs$treatment <- as.factor(bprs$treatment)
bprs$week <- as.integer(bprs$week)

#First 10 rows
kable(head(bprs),n=10) %>%
  kable_styling(full_width=F)

summary(bprs)

#make a backup id

bprs$subject <- as.integer(bprs$subject)

bprs$id2 <- 
  ifelse(bprs$treatment==1,
         bprs$subject,bprs$subject+20)

#table(bprs$id2,bprs$subject)

bprs$subject <- as.factor(bprs$subject)

library(wesanderson)
mypal <- wes_palette('Royal1')

bprs %>%
  ggplot(aes(x=week,
             y=bprs,
             group=id2,
             col=treatment)) +
  geom_line(size=1) +
  scale_colour_manual(values=mypal) +
  theme_minimal()

```

```{r}
#standard OLS model
mod1 <- lm(
  bprs~treatment + week,
  data=bprs)

summary(mod1)
```

```{r}
library(lme4)

# a random intercept model
modref <- lmer(
  bprs ~ treatment + week + (1 | id2),
  data = bprs, REML = FALSE)

summary(modref)


```

```{r}
library(lme4)

# a random intercept and 
#random slope model
modref2 <- lmer(
  bprs ~ treatment + week + (week | id2),
  data = bprs, REML = FALSE)

summary(modref2)

anova(modref,moderef2)


```